---
title: 'Health and Economic Consequences of Storms: 1950 - 2011'
author: "Charles Brown"
date: "Wednesday, March 18, 2015"
output:
  word_document: default
  pdf_document:
    toc: yes
  html_document:
    fig_caption: yes
    keep_md: yes
    toc: yes
---

### Synopsis  

This report explores the NOAA Strom Database and shows the types of Events that are most harmful to population health and that cause the most property and crop damage.

### Data Processing  

```{r, DefineURLAndDataDir}
strUrl <-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
strFilename <-"stormdata.bz2"
strDataDir <-""
```

* If needed, download compressed data [file](`r strUrl`).

```{r, DownloadData}
strPathFilename <-file.path(getwd(), strDataDir, strFilename, fsep = "/")
strPathFilename <-sub(pattern = "/{2,}", replacement = "/", x = strPathFilename)
if (!file.exists(strPathFilename))
{
    download.file(url = strUrl, destfile = strPathFilename, mode = "wb")   
}
```

* Read select columns into data frame `df`.

```{r, ReadDataIntoDataFrame}
df <-read.csv(file = strPathFilename, stringsAsFactors = FALSE, colClasses = c(rep("NULL",7),"character",rep("NULL",14),rep("numeric",3),"character","numeric","character",rep("NULL",9)))
summary(df)
```

* #### Calculate A New Event Group Variable `(df$EVTGROUP)` to summarize the `r length(table(df$EVTYPE))` event types.  
 
```{r, GroupEventTypes}
df$EVTGROUP <-NA
#
df[grep("^astro|^blow-out tide|^high tides",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Tides"
df[grep("^aval",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Avalanche"
df[grep("^blizz",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Blizzard"
df[grep("^coastal flood",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Coastal Flood"
df[grep("wind chill",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Wind Chill"
df[grep("debris flow",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Debris Flow"
df[grep("dense fog",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Dense Fog"
df[grep("smoke",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Dense Smoke"
df[grep("drought",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Drought"
df[grep("dust devil",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Dust Devil"
df[grep("^dust storm",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Dust Storm"
df[grep("^excessive heat|record/excessive heat",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Excessive Heat"
df[grep("extreme cold",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Extreme Cold/Wind Chill"
df[grep("^flood|flooding$",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Flood"
df[grep("^flash flood",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Flash Flood"
df[grep("frost",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Forst/Freeze"
df[grep("funnel",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Funnel Cloud"
df[grep("freezing fog",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Freezing Fog"
df[grep("hail",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Hail"
df[grep("^excessive heat|^extreme heat|^record heat|^heat|excessive heat$",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Heat"
df[grep("^heavy rain",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Heavy Rain"
df[grep("^heavy snow|snow and heavy snow|snow/heavy snow",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Heavy Snow"
df[grep("^high surf",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"High Surf"
df[grep("^high wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"High Wind"
df[grep("^hurricane|^typhoon",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Hurricane Typhoon"
df[grep("^ice storm|glaze/ice storm|sleet/ice storm|snow and ice storm|snow/ice storm",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Ice Storm"
df[grep("lake-effect snow",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Lake-Effect Snow"
df[grep("lakeshore flood",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Lakeshore Flood"
df[grep("^lightning",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Lightning"
df[grep("marine hail",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Marine Hail"
df[grep("marine high wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Marine High Wind"
df[grep("marine strong wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Marine Strong Wind"
df[grep("marine thunderstorm wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Marine Thunderstorm Wind"
df[grep("rip current",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Rip Current"
df[grep("seiche",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Seiche"
df[grep("sleet|^freezing drizzle|^freezing rain",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Sleet"
df[grep("storm surge",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Storm Surge"
df[grep("^strong wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Strong Wind"
df[grep("gusty thunderstorm wind|^severe thunderstorm wind|^thunderstorm wind",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Thunderstorm Wind"
df[grep("tornado",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Tornado"
df[grep("tropical depression",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Tropical Depression"
df[grep("tropical storm",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Tropical Storm"
df[grep("tsunami",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Tsunami"
df[grep("volcanic",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Volcanic Ash"
df[grep("^waterspout",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Waterspout"
df[grep("wildfire",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Wildfire"
df[grep("^winter storm",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"WinterStorm"
df[grep("winter weather",df$EVTYPE, ignore.case = TRUE),"EVTGROUP"] <-"Winter Weather"

```

* #### Calculate New Damage Amount Variables `(df$PROPDMGAMT & df$CROPDMGAMT)` for each event type.

```{r, CalculateDamageAmounts}
df$PROPDMGAMT <-NA
#
df[grep("[0-9]",df$PROPDMGEXP),"PROPDMGAMT"] <-(10^as.numeric(df[grep("[0-9]",df$PROPDMGEXP,),"PROPDMGEXP"])) * df[grep("[0-9]",df$PROPDMGEXP,),"PROPDMG"] 
df[grep("B",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMGAMT"] <-df[grep("B",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMG"] * 1000000000 
df[grep("M",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMGAMT"] <-df[grep("M",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMG"] * 1000000
df[grep("K",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMGAMT"] <-df[grep("K",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMG"] * 1000
df[grep("H",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMGAMT"] <-df[grep("H",df$PROPDMGEXP, ignore.case = TRUE),"PROPDMG"] * 100
#
df$CROPDMGAMT <-NA
#
df[grep("[0-9]",df$CROPDMGEXP),"CROPDMGAMT"] <-(10^as.numeric(df[grep("[0-9]",df$CROPDMGEXP,),"CROPDMGEXP"])) * df[grep("[0-9]",df$CROPDMGEXP,),"CROPDMG"] 
df[grep("B",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMGAMT"] <-df[grep("B",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMG"] * 1000000000 
df[grep("M",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMGAMT"] <-df[grep("M",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMG"] * 1000000
df[grep("K",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMGAMT"] <-df[grep("K",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMG"] * 1000
df[grep("H",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMGAMT"] <-df[grep("H",df$CROPDMGEXP, ignore.case = TRUE),"CROPDMG"] * 100
```

* #### Calculate New Total Vectors for Fatalities, Injuries, Property Damage and Crop Damage by Event Group `(df$EVTGROUP)`.

```{r, CalculateGroupTotals}
Fatalities <-tapply(X = df$FATALITIES, INDEX = df$EVTGROUP, FUN = sum, na.rm = TRUE)
Injuries <-tapply(X = df$INJURIES, INDEX = df$EVTGROUP, FUN = sum, na.rm = TRUE)
PropDmgAmt <-tapply(X = df$PROPDMGAMT, INDEX = df$EVTGROUP, FUN = sum, na.rm = TRUE)
CropDmgAmt <-tapply(X = df$CROPDMGAMT, INDEX = df$EVTGROUP, FUN = sum, na.rm = TRUE)
```

### Results 

* #### Ranking of Event Groups by Population Fatalities.

```{r, RankFatalitiesEvents}
(Fatalities <-sort(x = Fatalities, decreasing = TRUE))
```

* #### Ranking of Event Groups by Population Injuries. 

```{r, RankInjuriesEvents}
(Injuries <-sort(x = Injuries, decreasing = TRUE))
```

* #### Plot Ranking of Event Groups by Fatalities and Injuries. 

```{r, PlotHealthEffectResults}
par(mfrow = c(2,1))
barplot(Fatalities[1:5], ylab = "Total Fatalities")
barplot(Injuries[1:5], ylab = "Total Injuries")
par(mfrow = c(1,1))
```

* #### Ranking of Event Groups by Property & Crop Damage.   

```{r, RankDamageEventResults}
(PropDmgAmt <-sort(x = PropDmgAmt, decreasing = TRUE))
(CropDmgAmt <-sort(x = CropDmgAmt, decreasing = TRUE))
```

* #### Plot Ranking of Event Groups by Property & Crop Damage. 

```{r, PlotDamageEventResults}
barplot(PropDmgAmt[1:5]/1000000000, ylab = "Total Property Damage - $Billons")
barplot(CropDmgAmt[1:5]/1000000000, ylab = "Total Crop Damage - $Billions")
```


